Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2014-10-28 draft.tex
Rewrite recipe to determine acceptance effects

% Our method is closely related to Pearson's method of moments
% \citeneeded{}. One draws samples from a PDF $P(x | \nuvec)$ that
% depends on parameters $\nuvec$. Then one approximates the population
% (central) moments $x^k$ by the sample moments
Our method is an extension of the classical method of moments with
orthogonal functions~\cite[sec. 8.2]{James:2006zz}. The only difference
is that conventionally the angular functions are assumed
\emph{self-dual}, $\dual{f}_i = f_i$. However, it suffices that the
system of angular functions $\{f_i(\thvec)\}$ \emph{can} be transformed into
an orthonormal basis. It turns out to be much more convenient to work
in the basis natural to each process but the statistical properties
are equivalent for both approaches provided that one replace $f_i \to
\dual{f}_i$ appropriately. Using the ansatz
\begin{equation}
  \label{eq:dual-ansatz}
  \dual{f}_i = \sum_{j} a_{ij} f_j \,,
\end{equation}
the dual basis needs to be worked out case by case through solving the
linear system of equations (\ref{eq:def-ortho-rel}). For a selection
of hadron decays with a $b$ quark in the initial state and two leptons
in the final state, we list the dual bases in a series of appendices
\ref{app:btokll} through \ref{app:lambdabtolambdall}. Note that a
similar analysis was done in~\cite{Dighe:1998vk} for the decays $B \to
J/\psi \phi$ and $B \to J/\psi K^{*}$.

@@

The determination of a detector's angular acceptance is generally a difficult
task. Here, we will show a systematic method to estimate its effects
based on both the method of moments and MC simulations of the
detector.\\

In the ideal case, one would have an explicit probabilistic model of
the detector acceptance effects and could thus write down the full
forward model from which the measured events arise. In practice, this
would require much more CPU time than available, and one is therefore
forced to simplify the model. The standard approach is to generate
\emph{true} particle events $\lbrace x_t^{(k)}\rbrace =
\lbrace(\vec{\nu}^{(k)}_t,\vec\theta^{(k)}_t)\rbrace$ from a PDF
assumed to describe the bare physical process, and to propagate those
particles through a detailed simulation of the detector. The
observable traces that the particles leave in the detector are fed
into reconstruction algorithms resulting in the \emph{detector} events
$\lbrace x^{(k)}_d\rbrace = \lbrace(\vec{\nu}^{(k)}_d,
\vec\theta^{(k)}_d)\rbrace$.\fred{What about uncertainty in $\vec{\nu}, \vec{\theta}$? How does it enter the experimental likelihood? And how our method?}\\

As before, we assume the PDF $P$ can be decomposed into $N$ parts with angular observables $\lbrace S_k\rbrace$,
\begin{equation}
    P \equiv P(\vec\nu,\vec\theta | \lbrace S_k \rbrace)\,.
\end{equation}
We denote the estimator of an
observable $S_k$ as $\est{S_k}$, which is affected by the detector acceptance. We assume now that there is a linear relation
between the true angular observables $\lbrace S_k^\text{true}\rbrace$, and the estimators $\lbrace \est{S_k}\rbrace$, \fred{We don't need the $^{\text true}$.}
\begin{equation}
    \label{eq:linear-assumption}
    \est{S_i} = \sum_{j = 1}^N M_{i j} S_j^\text{true}\,.
\end{equation}
This assumption is based on a smooth angular acceptance
function. \fred{Check the LHCb \href{http://inspirehep.net/record/1287929}{paper} on $B \to K \ell \ell$: their
  acceptance (fig. 1) is not smooth due to a cut!} \fred{I don't understand. Need a derivation of why this hinges on smoothness} While this will not be true for
the laboratory frame, smoothness may be recovered in a boosted frame,
e.g. the rest frame of the decaying particle.  In that case, the
matrix $M_{ij}$ can be fully determined through the following steps.
\begin{enumerate}
    \item For fixed $j$, let $S_j^\text{true} = 1$, and $S_k^\text{true} = 0\,\forall k\neq j$.
    \item Draw simulated events $\lbrace x_t^{(k)}\rbrace =
      \lbrace(\vec{\nu}^{(k)}_t,\vec\theta^{(k)}_t)\rbrace$ \fred{don't use $k$ as superscript and subscript} from the PDF $P_j$,
      \begin{equation}
        \label{eq:single-comp}
            P_j \equiv P(\vec\nu,\vec\theta | \lbrace S_j = 1, S_{k\neq j} = 0\rbrace)\,.
        \end{equation}
    \item Use the true events as input for the MC simulation of the detector in order to obtain the detector
        events $\lbrace x^{(k)}_d\rbrace = \lbrace(\vec{\nu}^{(k)}_d,\vec\theta^{(k)}_d)\rbrace$.
    \item Determine the uncorrected observables $\lbrace \est{S_k} \rbrace$ based on the detector
        events $\lbrace x^{(k)}_d\rbrace$. For all intents and purposes, automation of this step is only feasible
        with to the method of moments, since it does not suffer form convergence problems.
    \item By virtue of \refeq{linear-assumption} and step 1, the $j$-th column of $M$ has been determined via
        \begin{equation}
            M_{ij} = \est{S_i}\,.
        \end{equation}
    \item Repeat the above steps for all values of $j$ to determine all columns of $M$.
\end{enumerate}

It remains to invert the relation \refeq{linear-assumption}, so that one can determine the unfolded
angular observables $\lbrace \wwhat{S_k}\rbrace$ as
\begin{equation}
    \wwhat{S_i} = \danny{N_\eps} \left(M^{-1}\right)_{ij} \est{S_j}\,.
  \end{equation}
  \fred{What happens to the uncertainties? Unfolding is used in many
    places, so there are elaborate recipes. Should we assume the same
    uncertainties as before? Since it is linear, rescaling should be sufficient. Or compute covariance directly for transformed observables} \danny{Here $N_\eps$ is a normalization factor
    based on the detector acceptance function $\eps$. It emerges since
    the product $P(x) \eps(x)$ is not a PDF; it is not normalized
    appropriately\footnote{We thank X.X\danny{Marcin: Please insert
        whatshisname} for pointing this out.}. We can determine
    $N_\eps$ by imposing
    \begin{gather}
        1 \equiv \int \rmdx{x} P(x | \lbrace \wwhat{S}_k\rbrace) = N_\eps \sum_{k,j} \left(M^{-1}\right)_{kj} \est{S_j}\int \rmdx{x} f_k(x)\,,\\
        \Rightarrow N_\eps \equiv \frac{1}{\sum_{k,j} \left(M^{-1}\right)_{kj} \est{S_j} \int \rmdx{x} f_k(x)}\,.
    \end{gather}
}

We emphasize that this method hinges on the assumption in \refeq{linear-assumption}\danny{which is well motivated:
the PDF is a linear combination of the basis functions $f_i$, and the moments are therefore a linear
functional of said basis functions. Nevertheless,}
we thus suggest to explicitly
test the linearity. One way of testing involves repeating the previous procedure
with $S_j^\text{true} = 1/z$, $z > 0$. If the linearity assumption holds, one should find
\begin{equation}
    M_{ij}(z) = z \est{S_i}\,,
\end{equation}
for a fixed value of $j$. \fred{This is useless. \refeq{single-comp}
  says: draw from a single component $f_j$. This $f_j$ defines the
  distribution, and whatever factor you put in front of it, you get
  the same distribution of events.}

\danny{In order to better understand the detector aceptance effects, we propose two steps. For the sake of
    clarity of our argument, we assume in the following that the PDF $P$ is expanded in products of
    appropriately chosen Legendre polynomials to some multi-index order $\vec{l}_P^{(0)}$. Similarly,
    moments and detector effects are nominally determined up to some order $\vec{l}_M^{(0)}$, with
    $\vec{l}_M^{(0)} \leq \vec{l}_P^{(0)}$. Here and in the following all comparisons are to be read
    for each compoenent.\\
    For the first step, we remind that the unfolding matrix $M$ depends implicitly on the physical
    PDF $P$. We therefore recommend variation of the order of the expansion of $P$ to $\vec{l}_P \geq \vec{l}_P^{(0)}$
    component-wise. This way one can estimate the mixing of PDF terms of order $\vec{l}_P$ into moments of order $\vec{l}_M^{(0)}$
    or lower.\\
    In a second step, we assume using the nominal expansion of the PDF up to order $\vec{l}_P^{(0)}$.
    With moments of order $\vec{l}_M \geq \vec{l}_M^{(0)}$ one can therefore find the impact of the
    acceptance effects on the mixing of moments of order $\vec{l}_M^{(0)}$ and below
    into moments beyond the order of the PDF.
    Our recommendation is not illustrated at the hand of numerical examples, which would necessitate
    access to realistic detector models. The latter are not available to persons outside the
    respective collaborations. The proposed studies are therefore beyond the scope of this work.
}

\fred{I'm puzzled. If you can do the calculation to some order
  $\vec{l}_P$, why would you use a lower order for the actual result?
  I.e., if you have the NNLO results, why would you limit yourself to
  NLO?}

@@

The PDF for the decay $\bar{B}\to\bar{K}\ell^+\ell^-$ has been calculated for the most
complete basis of dimension-six $b\to s \ell^+\ell^-$ operators. It reads \cite{Bobeth:2007dw,Bobeth:2012vn}
\begin{equation}
    P(q^2, \cos\theta_1) = \frac{1}{\Gamma} \frac{\rmdx{^4\Gamma}}{\rmdx{q^2} \rmdx{\cos\theta_1}} = \frac{a(q^2)}{\Gamma} + \frac{b(q^2)}{\Gamma} \cos\theta_1 + \frac{c(q^2)}{\Gamma} \cos^2\theta_1 = \sum_i P_i f_i(\cos\theta_1)\,,
\end{equation}
where the basis of angular functions $\lbrace f_i\rbrace$ reads
\begin{equation}
\begin{aligned}
    f_1 & = 1\,, &
    f_2 & = \cos\theta_1\,, &
    f_2 & = \cos^2\theta_1\,.
\end{aligned}
\end{equation}
Here we denote the dilepton mass squared as $q^2$, and the dilepton helicity angle as $\theta_1 \equiv \theta_{\ell}$. We transform the PDF from a $\cos\theta_1$-dependence
to a $\theta_1$-dependence,
\begin{equation}
    P(q^2, \theta_1) = \sum_i P_i f_i(\cos\theta_1) \sin \theta_1\,.
\end{equation}
where we define $P_{\lbrace 1,2,3\rbrace} \equiv \lbrace a, b, c\rbrace / \Gamma$.\\

The dual basis $\lbrace \dual{f}_i\rbrace$ reads
\begin{equation}
\begin{aligned}
    \dual{f}_1 & = \frac{9}{8} - \frac{15}{8}\cos^2\theta_1\,, &
    \dual{f}_2 & = \frac{3}{2}\cos\theta_1\,, &
    \dual{f}_3 & = -\frac{15}{8} + \frac{45}{8}\cos^2\theta_1\,,
\end{aligned}
\end{equation}
such that
\begin{equation}
    \int_0^\pi \rmdx{\theta_1} \dual{f}_i(\theta_1) P(q^2, \theta_1) = S_i(q^2)\,.
\end{equation}

@@

>>>>

% Our method is closely related to Pearson's method of moments
% \citeneeded{}. One draws samples from a PDF $P(x | \nuvec)$ that
% depends on parameters $\nuvec$. Then one approximates the population
% (central) moments $x^k$ by the sample moments
Our method is an extension of the classical method of moments with
orthogonal functions~\cite[sec. 8.2]{James:2006zz}. The only difference
is that conventionally the angular functions are assumed
\emph{self-dual}, $\dual{f}_i = f_i$. However, it suffices that the
system of angular functions $\{f_i(\thvec)\}$ \emph{can} be transformed into
an orthonormal basis. It turns out to be much more convenient to work
in the basis of Legendre polynomials but the statistical properties
are equivalent for both approaches provided that one replace $f_i \to
\dual{f}_i$ appropriately. Using the ansatz
\begin{equation}
  \label{eq:dual-ansatz}
  \dual{f}_i = \sum_{j} a_{ij} f_j \,,
\end{equation}
the dual basis needs to be worked out case by case through solving the
linear system of equations (\ref{eq:def-ortho-rel}). For a selection
of hadron decays with a $b$ quark in the initial state and two leptons
in the final state, we list the dual bases in a series of appendices
\ref{app:btokll} through \ref{app:lambdabtolambdall}. Note that a
similar analysis was done in~\cite{Dighe:1998vk} for the decays $B \to
J/\psi \phi$ and $B \to J/\psi K^{*}$.

@@

The determination of a detector's performance to detect signal events
with accurate determination of the event's angles is generally a difficult
task. In the ideal case, one would have an explicit probabilistic model of
the detector acceptance effects and could thus write down the full
forward model from which the measured events arise. In practice, this
would require much more CPU time than available, and one is therefore
forced to simplify the model. The standard approach is to generate
the \emph{true} particle events $\lbrace x_\text{t}^{(n)}\rbrace =
\lbrace(\vec{\nu}^{(n)}_\text{t},\vec\theta^{(n)}_\text{t})\rbrace$, $n=1,\dots,N_\text{t}$, from a PDF
assumed to describe the bare physical process, and to propagate those
particles through a detailed simulation of the detector.
The observable traces that the particles leave in the detector are fed
into reconstruction algorithms resulting in the \emph{detector} events
$\lbrace x^{(n)}_\text{d}\rbrace = \lbrace(\vec{\nu}^{(n)}_\text{d},
\vec\theta^{(n)}_\text{d})\rbrace$, $n=1,\dots,N_\text{d}$.
Most generally, the detected events $\lbrace x_\text{d}^{(k)}\rbrace$
are distributed as $P_\text{d}$,
\begin{equation}
    P_\text{d}(x_\text{d}) = \int P_\text{t}(x_\text{t}) E(x_\text{d}, x_\text{t}) \rmdx{x_\text{t}}\,.
\end{equation}
Here $P$ is the probability distribution of the true events, and the kernel $E(x_\text{d}, x_\text{t})$
parametrizes acceptance and resolution
effects. Note that $P_\text{d}$ is \emph{not} a PDF, since it is not normalized,
\begin{equation}
    \int P_\text{d}(x) \rmdx{x} \neq 1\,.
\end{equation}
For the sake of clarity, we will restrict ourselves
to pure acceptance effects and suppress resolution effects in the remainder of this section; i.e.,
\begin{equation}
    E(x_\text{d}, x_\text{t}) \equiv \eps(x_\text{t}) \delta(x_\text{d}, x_\text{t})\,.
\end{equation}
We will refer to $\eps$ as the \emph{detector acceptance} function. In what follows,
we propose a systematic methods to unfold all effects of $\eps$ through MC simulations
and the method of moments.\\

We will now proceed with the explicit usage of a uniangular PDF with $x = \cos\theta$.\footnote{%
    The generalization of this section to multiangular PDFs is straighforward. It can be achieved
    by promoting $x$ to a vector, promoting the Legendre polynomials to products of independent polynomials
    and the indices $i,j,k,m$ to multi-indices.
}
Let us define the PDF in terms of the Legendre polymials (i.e., $f_k(x) \equiv P_k(x)$),
\begin{equation}
    P(x) \equiv \sum_k S_k f_k(x),
\end{equation}
where $k = 0, 1, \dots$ denotes the total angular momentum of the observables.
From the normalization of the PDF then follows $S_0 = 1/2$. From positivity of $P$ one
obtains $|S_k| \leq 1/2$ for angular momentum $k > 0$. More stringent relations between the
$S_k$ might hold, but are beyond the scope of this work. For latter use, we define
$\vec{S}^{(m)}_k \equiv 1/2 (\delta_{k,0} + \delta_{k,m})$, and we emphasize that
\begin{equation}
    P^{(m)}(x) \equiv P(x | \lbrace S^{(m)}_k \rbrace)
\end{equation}
is a PDF. The dual basis of angular functions follows then from the normalization
and orthogonality of the Legendre polynomials, and one therefore has
\begin{equation}
    \tilde{f}_k(x) = \frac{2 k + 1}{2} f_k(x),\qquad \int_{-1}^{+1} \tilde{f}_k(x) f_l(x) \rmdx{x} = \delta_{k,l}\,.
\end{equation}

We now define the \emph{simulated raw} moments $\lbrace Q_{i}^{(m)}\rbrace$,
\begin{equation}
    Q_i^{(m)} \equiv \int_{-1}^{+1} \tilde{f}_i(x) P^{(m)}(x) \eps(x) \rmdx{x} = \frac{1}{2} \int_{-1}^{+1} \tilde{f}_i(x) \big[f_0(x) + f_m(x)\big] \eps(x) \rmdx{x}\,,
\end{equation}
which are instrumental to our recipe. MC estimators for these simulated moments can be constructed from specifically crafted detector events $x_\text{d}^{(n,m)}$, $n = 1, \dots, N_\text{d}^{(m)}$, where
\begin{equation}
    x_\text{d}^{(n,m)} \sim P_\eps^{(m)} \equiv \frac{P^{(m)} \eps}{R^{(m)}}\,.
\end{equation}
Here the normalization $R^{(m)}$ is chosen such that $\int P_\eps^{(m)}(x) \rmdx{x} = 1$.
We emphasize that $R^{(m)}$ can estimated as $\est{R}^{(m)} = N_\text{d}^{(m)} / N_\text{t}$, where $N_\text{t}$ corresponds to the number of simulated true events.
The estimators then read
\begin{equation}
    \est{Q}_i^{(m)} \equiv \est{R}^{(m)} \frac{1}{N_\text{d}} \sum_n^{N_\text{d}} \tilde{f}_i(x_\text{d}^{(n,m)})
\end{equation}

Linearity of the integral over $x$ and convergence of the expansion of $P_\eps^{(m)}$ in terms Legendre polynomials ensures that
\begin{equation}
    Q_i^{(m)} = \sum_j M_{ij} S_j^{(m)}\,.
\end{equation}
We call the matrix $M_{ij}$ the \emph{unfolding matrix}, which is specific to our decay at hand. Given our definition of
$S_k^{(m)}$ it can easily be calculated as
\begin{equation}
    M_{ij} = \begin{cases}
        2 Q_i^{(0)}                          & j = 0\,,\\
        2\left(Q_i^{(j)} - Q_i^{(0)}\right)  & j \neq 0\,,\\
    \end{cases}
\end{equation}
and its MC estimator $\est{M}_{ij}$  can be obtained through the replacements $Q_i^{(m)} \to \est{Q}_i^{(m)}$.\\

In order to finally extract the angular observables from data, we use the \emph{measured raw moments}. Their MC estimator
$\lbrace \est{Q}_i\rbrace$ -- based on the \emph{detected events} $x^{(n)}$, $n=1,\dots,N$ -- reads
\begin{equation}
    \est{Q}_i \equiv R \frac{1}{N} \sum_n^N \tilde{f}_i(x^{(n)})\,.
\end{equation}
Here $R \equiv \int P(x) \eps(x)$ is an a priori unknown normalization that also occurs in the determination of the branching
ratio of the specific decay. We then obtain MC estimators of the angular observables via
\begin{equation}
    \est{S}_i \equiv \sum_{i} \left(\hat{M}^{-1}\right)_{ij} \est{Q}_j\,.
\end{equation}
The precise value of $R$ depends on the a priori unknown angular observables $\lbrace S_i\rbrace$.
However, ratios and other suitable combinations of the estimators $\lbrace \est{S}_i\rbrace$ are independent
of $R$.

\todo{size of $M$ in relation to the maximum angular momentum of $\eps$}

@@

The PDF for the decay $\bar{B}\to\bar{K}\ell^+\ell^-$ has been calculated for the most
complete basis of dimension-six $b\to s \ell^+\ell^-$ operators. It reads \cite{Bobeth:2007dw,Bobeth:2012vn}
\begin{equation}
    P(q^2, \cos\theta_1)
        = \frac{1}{\rmdx{\Gamma}/\rmdx{q^2}} \frac{\rmdx{^2\Gamma}}{\rmdx{q^2} \rmdx{\cos\theta_1}}
        = \frac{a(q^2)}{\rmdx{\Gamma}/\rmdx{q^2}} + \frac{b(q^2)}{\rmdx{\Gamma}/\rmdx{q^2}} \cos\theta_1 + \frac{c(q^2)}{\rmdx{\Gamma}/\rmdx{q^2}} \cos^2\theta_1
        = \sum_i S_i p_i(\cos\theta_1)\,,
\end{equation}
with the conventional observables $a(q^2)$ through $c(q^2)$, and $\rmdx{\Gamma}/\rmdx{q^2} = 2a + 2/3 c$. We conveniently use the Legendre polynomials
$p_i(x)$, $i=0,1,2$,
\begin{equation}
\begin{aligned}
    p_0(x) & = 1\,, &
    p_1(x) & = x\,, &
    p_2(x) & = \frac{1}{2} (3x^2 - 1)\,,
\end{aligned}
\end{equation}
as our basis of angular functions. Our basis of angular observables then
translates to the conventional basis as
\begin{equation}
\begin{aligned}
    S_0 & = \frac{1}{2}\,, &
    S_1 & = \frac{b}{\Gamma}\,, &
    S_2 & = \frac{2c}{3\Gamma}\,.
\end{aligned}
\end{equation}
In this case, the dual basis is simply given by $\tilde{p}_i(x) = (2 i + 1)/2 p_i(x)$
such that
\begin{equation}
    \int_0^\pi \rmdx{\theta_1} \dual{p}_i(\cos\theta_1) P(q^2, \cos\theta_1)\sin\theta_1 = S_i(q^2)\,.
\end{equation}

@@

\todo{rewrite to use Legendre polynomials}
